---
title: "Demonstrate isofemales"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demonstrate isofemales}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 6)
```

# Creation of isofemale lines
The package GenomeAdmixR was designed to simulate the generation of iso-female 
lines, and simulate genomic changes during, and after, formation of iso-female 
lines. Furthermore, it allows for simulation of effects after crossing 
individuals from iso-female lines.

```{r load libraries}
library(GenomeAdmixR)
library(ggplot2)
packageVersion("GenomeAdmixR")
```
We will analyze whether we can genetically tell apart two populations, where one 
population was founded by crossing two isofemale lines that were created from 
the same source population, and the other population was founded by crossing 
two isofemale lines created from different source populations. 

First, we generate two source populations using \code{simulate_admixture} and 
then increase the founder labels in the second population to ensure that both 
populations do not share founder labels.

```{r create populations}
pops <-
  simulate_admixture_migration(pop_size = c(100, 100),
                               initial_frequencies =
                                 list(c(1, 1, 1, 1, 0, 0, 0, 0),
                                      c(0, 0, 0, 0, 1, 1, 1, 1)),
                               total_runtime = 1000,
                               migration_rate = 0)

pop_1 <- pops$population_1
pop_2 <- pops$population_2
```

To ensure that our two wild populations are not inbred, we estimate the average 
Linkage Disequilibrium (LD) for both populations using the function 
calculate_LD. In the absence of inbreeding, we expect the average LD to be low 
or zero.

```{r calculate LD}
mean(calculate_ld(pop = pop_1,
                  sampled_individuals = 10,
                  number_of_markers = 30,
                  random_markers = TRUE)$ld_matrix,
     na.rm = TRUE)
mean(calculate_ld(pop = pop_2,
                  sampled_individuals = 10,
                  number_of_markers = 30,
                  random_markers = TRUE)$ld_matrix,
     na.rm = TRUE)
```
Now we can start to generate isofemale lines from both populations, using \code{create_iso_female}.
```{r create isofemale lines}
iso_females_pop_1 <- create_iso_female(pop_1,
                                       n = 2,
                                       inbreeding_pop_size = 100)
iso_females_pop_2 <- create_iso_female(pop_2,
                                       n = 2,
                                       inbreeding_pop_size = 100)
```
Using the function \code{create_population_from_individuals} we then create
three populations, two where the isofemales are drawn from the same source 
population, and one where we mix the two, e.g. 1x1, 1x2 and 2x2, where 
isofemales 1 and 2 indicate isofemales from source populations 1 and 2 
respectively. 
```{r create from individuals}
pop_1_1 <- simulate_admixture(iso_females_pop_1,
                              pop_size = 1000,
                              total_runtime = 1000,
                              morgan = 1)

pop_1_2 <- simulate_admixture(list(iso_females_pop_1[[1]],
                                   iso_females_pop_2[[1]]),
                               pop_size = 1000,
                               total_runtime = 1000,
                               morgan = 1)

pop_2_2 <- simulate_admixture(iso_females_pop_2,
                              pop_size = 1000,
                              total_runtime = 1000,
                              morgan = 1)
```
Then, using the function \code{calculate_fst} we calculate all pairwise FST 
values. Here, we expect to find the highest FST for the 1x2 comparison. 
```{r FST calculation}
f1 <- calculate_fst(pop_1_1, pop_1_2,
                    sampled_individuals = 10, number_of_markers = 100)
# this one should be highest
f2 <- calculate_fst(pop_1_1, pop_2_2,
                    sampled_individuals = 10, number_of_markers = 100)
f3 <- calculate_fst(pop_1_2, pop_2_2,
                    sampled_individuals = 10, number_of_markers = 100)
f1
f2
f3
```
Which confirms our hypothesis and demonstrates how combining isofemale lines 
from different source populations generates a population with maximal genetic 
diversity. 
