---
title: "Visualization"
author: "Thijs Janzen"
vignette: >
  %\VignetteIndexEntry{Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
  
```{r setup, include=FALSE}
library(GenomeAdmixR)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 6)
```

# Visualization

The GenomeAdmixR package provides many different visualization options, here we will explore a variety of them.
First, we simulate a scenario with selection, to obtain somewhat meaningful results.

```{r simulate}
select_matrix <- matrix(nrow = 1, ncol = 5)
select_matrix[1, ] <- c(0.5, 1, 1 + 0.05, 1 + 0.1, 0)
population <- simulate_admixture(pop_size = 1000,
                                 number_of_founders = 2,
                                 total_runtime = 200,
                                 select_matrix = select_matrix,
                                 markers = c(0.5, seq(0, 1, length.out = 100)))
```

Now, we can first view whether selection on our marker has yielded an increase in frequency:

## Plot over time

```{r plot over time}
plot_over_time(population$frequencies, focal_location = 0.500)
```

Indeed, we observe that over time the frequency of the allele under selection (0) increases to fixation due to selection. 

## Plot Frequencies
How are the alleles scattered across the genome? we can answer that with the function plot_frequencies:
```{r plot frequencies}
plot_frequencies(population, locations = seq(0, 1, length.out = 1000))
```
As expected, we observe a huge increase around the location of the marker under selection (at 0.5 Morgan).

## plot difference frequencies
If instead, we are interested in the change in frequency of a marker, we can do so using plot_difference_frequencies.

```{r plot difference frequencies}
plot_difference_frequencies(population)
```

## plot_start_end
Visualized in another way, plot_start_end plots the frequency distributions at the start and at the end of the simulation:
```{r plot start end}
plot_start_end(population)
```

## joyplots
If, indeed, we are more interested in the progression over time, we can also consult a so called 'joyplot', or 'ridgeplot':
```{r joyplot}
plot_joyplot_frequencies(population$frequencies,
                         time_points = c(0, 10, 25, 50, 100, 199))
```

## Plotting individual chromosomes
Lastly, individual chromosomes can be visualized using the standard plotting functions, where both the entire chromosome, and a fraction of the chromosome, can be visualized:

```{r plot chromosome}
plot(population$population[[1]])

plot_chromosome(population$population[[1]]$chromosome1, xmin = 0.45, xmax = 0.55)
```
